---
title: "Multiple linear regression: Interactions"
subtitle: "Lecture 4"  
author: 
  - "Manuel Villarreal"
date: '08/28/24'
output:
  xaringan::moon_reader:
    self_contained: true
    css: ['xaringan-themer.css', 'set-colors.css']
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.width = 7,
  fig.height = 5.8,
  fig.align = 'center',
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, echo=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#005f86",
  secondary_color = "#f8971f",
  inverse_header_color = "#FFFFFF"
)
```

```{r xaringan-extra, echo=FALSE, warning=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ))
rmarkdown::html_dependency_font_awesome()
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE)
```

```{r data-sim-example, echo = FALSE}
library(flipbookr)
set.seed(291792)
# simulating data for experiment one
blood <- data.frame("id" = seq(from = 1, to = 200),
                    "sex" = rep(x = c("female", "male"), each = 100),
                    "age" = round(
                      x = rnorm(n = 200, mean = 45, sd = 3) + 
                        rexp(n = 200, rate = 1/4),
                      digits = 0),
                    "height" = round(
                      x = rnorm(n = 200, 
                                mean = rep(x = c(170.7, 179.8), each = 100),
                                sd = rep(x = c(6.35, 7.62), each = 100)), 
                      digits = 0)) |>
  dplyr::mutate("blood_pressure" = round(x = 
                  96.6 + 16 * as.numeric(sex == "male") + 
                  0.7 * age - 0.4 * age * as.numeric(sex == "male") +
                  rnorm(n = 200, mean = 0, sd = 3), digits = 1))

mbp <- lm(formula = blood_pressure ~ age + sex, data = blood)
```

### Recap

- In the last class we ran our first multiple linear regression using age and 
sex as predictors of blood pressure.

--

- Adding the sex variable to our model made the residuals look better when 
plotted as a function of the observation order (id) which was not the case 
for the simple linear model.

--

- However, we still need to see that the residuals behave as expected (mean 0 
and constant variance) as a function of other variables.

---
### Residuals as a function of age

.pull-left[
```{r residual-age, eval = FALSE}
# Plot residual against
# its corresponding id (order)
plot(x = blood$age, y = mbp$residuals, ann = FALSE, 
     axes = FALSE, cex = 2, pch = 21,
     col = "white", bg = "#f8971f")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
axis(side = 2, cex.axis = 1.5, las = 2)
mtext(text = "Residuals", side = 2, line = 2.5,
      cex = 2)
mtext(text = "age", side = 1, line = 3,
      cex = 2)

# what should we expect?
abline(h = 0, col = "#005f86", lty = 4, 
       lwd = 3.5)
```
]

.pull-right[
```{r residual-age-out, ref.label = "residual-age", fig.align = 'center', echo = FALSE, fig.height = 6}

```
]

---
### Residuals of a function of age

.panelset[
  .panel[.panel-name[Residuals]
  
- When plotting the residuals as a function of age we can't see anything wrong 
immediately.

- However, we can still stratify our residuals by sex.
  ]
  .panel[.panel-name[Females]
.pull-left[
```{r residual-f-age, eval = FALSE}
par(mar = c(4.2,4.2,1,1))
plot(x = blood$age[which(blood$sex == "female")],
     y = mbp$residuals[which(blood$sex == "female")],
     xlim = c(37, 65), ylim = c(-10, 10),
     ann = FALSE, axes = FALSE, cex = 2, pch = 21,
     col = "white", bg = "#f8971f")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
axis(side = 2, cex.axis = 1.5, las = 2)
mtext(text = "Residuals", side = 2, line = 2.5,
      cex = 2)
mtext(text = "Age", side = 1, line = 3,
      cex = 2)
# what should we expect?
abline(h = 0, col = "#005f86", lty = 4, 
       lwd = 3.5)
```
]
.pull-right[
```{r residual-f-age-out, ref.label = "residual-f-age", fig.align = 'center', echo = FALSE, fig.height = 5}

```
]
]
  .panel[.panel-name[Males]
.pull-left[
```{r, residual-m-age, eval = FALSE}
par(mar = c(4.2,4.2,1,1))
plot(x = blood$age[which(blood$sex == "male")],
     y = mbp$residuals[which(blood$sex == "male")],
     xlim = c(37, 65), ylim = c(-10, 10),
     ann = FALSE, axes = FALSE, cex = 2, pch = 21,
     col = "white", bg = "#003f5c")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
axis(side = 2, cex.axis = 1.5, las = 2)
mtext(text = "Residuals", side = 2, line = 2.5,
      cex = 2)
mtext(text = "Age", side = 1, line = 3,
      cex = 2)
# what should we expect?
abline(h = 0, col = "#005f86", lty = 4, 
       lwd = 3.5)
```
]
.pull-right[
```{r residual-m-age-out, ref.label = "residual-m-age", fig.align = 'center', echo = FALSE, fig.height = 5}

```
]
  ]
]

---
### Stratified residuals

- Once we look at the residuals of the model independently for each of our two
groups we can see that there's still a small bias in our model.

--

- For example, we noticed that on average our residuals for the female group
tend to increase as a function of age.

--

- While the residuals for the male group tend to decrease with age. 

--

- These type of results suggest to us that there might be an interaction between
the two variables.

---
### Interactions

- From the previous plots we can see that, once we stratify the residuals by 
sex, we can see a trend in the residuals:

--

  - Residuals for female participants increase as a function of age.
  
--

  - Residuals for male participants decrease as a function of age.
  
--

- This suggests that there might be an interaction between sex and age that 
could be important for the model.

---
### Adding interactions to an MLR

- Formally, we can express the interaction model as:
$$\text{blood pressure}_i = \beta_0 + \beta_1\text{age}_i + \beta_2\text{sex}_i + \beta_3\text{age}_i \text{sex}_i + \epsilon_i$$

--

- In R we can add interactions to a linear model in two different ways:

```{r}

mlr_int_1 <- lm(formula = blood_pressure ~ age + sex + age * sex, 
                data = blood)
mlr_int_2 <- lm(formula = blood_pressure ~ age:sex, data = blood)
```

--

- Just like before, the output of the function includes useful elements that we
can use to evaluate the model.

---
### Adding interactions to an MLR

- An important thing to note is that the `formula` notation that includes the
colon:

```{r, eval=FALSE}
mlr_int_2 <- lm(formula = blood_pressure ~ age:sex, data = blood)
```

has a different interpretation of the estimated values in comparison to the 
notation that has the sums and products

```{r, eval=FALSE}
mlr_int_1 <- lm(formula = blood_pressure ~ age + sex + age * sex, 
                data = blood)
```

--

- The reason for this difference is that the model produced by the **colon**
notation model does not include a specific coefficient for age, and instead has 
one age coefficient (parameter) for each group.

---
### Residual plots

.panelset[
.panel[.panel-name[Interactions]
.pull-left[
```{r line-plot, eval=FALSE}
par(mar = c(4.2,4.2,1,1))
curve(expr = mbp$coefficients[1] + mbp$coefficients[2] * x,
      lwd = 2, from = 36, to = 69, ylim = c(115, 145),
      col = "#f8971f66", lty = 2, las = 1,
      xlab = "Age", ylab = "Blood pressure", cex = 1.5)
curve(expr = mlr_int_1$coefficients[1] + 
        mlr_int_1$coefficients[2] * x,
      lwd = 3, from = 36, to = 69, add = TRUE, 
      col = "#f8971f")
curve(expr = (mbp$coefficients[1] + mbp$coefficients[3]) + 
        mbp$coefficients[2] * x,
      lwd = 2, from = 36, to = 69, add = TRUE, 
      col = "#003f5c66", 
      lty = 2)
curve(expr =
        (mlr_int_1$coefficients[1] + mlr_int_1$coefficients[3]) +
        (mlr_int_1$coefficients[2] + mlr_int_1$coefficients[4]) * x,
      lwd = 3, from = 36, to = 69, add = TRUE, col = "#003f5c")
```
]
.pull-right[
```{r line-plot-out, ref.label = "line-plot", fig.align = 'center', echo = FALSE, fig.height = 5}

```
]
]
.panel[.panel-name[Histogram]
.pull-left[
```{r err-hist, eval = FALSE}
par(mar = c(4.2,4.2,1,1))
# Then we can draw the histogram
hist(x = mlr_int_2$residuals, breaks = 30,
     axes = FALSE, ann = FALSE, freq = FALSE,
     border = "white", col = "#f8971f")
abline(v = mean(mlr_int_1$residuals), 
       lty = 2, lwd = 3, col = "#005f86")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
mtext(side = 1, cex = 1.6, line = 3,
      text = expression(
        x = paste("Residuals: ",
                  hat(epsilon)[i])))
```
]
.pull-right[
```{r err-hist-out, ref.label = "err-hist", fig.align = 'center', echo = FALSE, fig.height = 5}

```
]
]
.panel[.panel-name[Females]
.pull-left[
```{r residual-f-int, eval = FALSE}
par(mar = c(4.2,4.2,1,1))
plot(x = blood$age[which(blood$sex == "female")],
     y = mlr_int_1$residuals[which(blood$sex == "female")],
     xlim = c(37, 65), ylim = c(-10, 10),
     ann = FALSE, axes = FALSE, cex = 2, pch = 21,
     col = "white", bg = "#f8971f")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
axis(side = 2, cex.axis = 1.5, las = 2)
mtext(text = "Residuals", side = 2, line = 2.5,
      cex = 2)
mtext(text = "Age", side = 1, line = 3,
      cex = 2)
# what should we expect?
abline(h = 0, col = "#005f86", lty = 4, 
       lwd = 3.5)
```
]
.pull-right[
```{r residual-f-int-out, ref.label = "residual-f-int", fig.align = 'center', echo = FALSE, fig.height = 5}

```
]
]
.panel[.panel-name[Males]
.pull-left[
```{r, residual-m-int, eval = FALSE}
par(mar = c(4.2,4.2,1,1))
plot(x = blood$age[which(blood$sex == "male")],
     y = mlr_int_1$residuals[which(blood$sex == "male")],
     xlim = c(37, 65), ylim = c(-10, 10),
     ann = FALSE, axes = FALSE, cex = 2, pch = 21,
     col = "white", bg = "#003f5c")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
axis(side = 2, cex.axis = 1.5, las = 2)
mtext(text = "Residuals", side = 2, line = 2.5,
      cex = 2)
mtext(text = "Age", side = 1, line = 3,
      cex = 2)
# what should we expect?
abline(h = 0, col = "#005f86", lty = 4, 
       lwd = 3.5)
```
]
.pull-right[
```{r residual-m-int-out, ref.label = "residual-m-int", fig.align = 'center', echo = FALSE, fig.height = 5}

```
]
]
.panel[.panel-name[Auto-correlation]
.pull-left[
```{r auto-cor-int, eval = FALSE}
# We can use the function built in r
acf(x = mlr_int_1$residuals, las = 1, lwd = 2,
    main = "Series: residuals", 
    ylab = "", 
    xlab = "Lag", cex.lab = 1.7, cex.main = 1.5,
    cex.axis = 1.3)
mtext(text = "Auto-correlation", side = 2, 
      line = 2.8, cex = 1.7)
```
]
.pull-right[
```{r auto-cor-int-out, ref.label = "auto-cor-int", fig.align = 'center', echo = FALSE, fig.height = 5}

```
]
]
]

---
### Conclusion

- Now that we have added interactions to the blood pressure model our estimated
residuals look good.

--

- Their mean is close to 0, they have constant variance (rectangle when plotted
against other variables), and they have a low auto-correlation which suggests 
independence.

--

- This seems to be a good model for our data but how can we know if it is?

--

- Furthermore, when do we decide to stop adding covariates (the $X$'s) to our
model?
