---
title: "Transformation of independent variables"
subtitle: "Lecture 6"  
author: 
  - "Manuel Villarreal"
date: '09/30/24'
output:
  xaringan::moon_reader:
    self_contained: true
    css: ['xaringan-themer.css', 'set-colors.css']
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.width = 7,
  fig.height = 5.8,
  fig.align = 'center',
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, echo=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#005f86",
  secondary_color = "#f8971f",
  inverse_header_color = "#FFFFFF"
)
```

```{r xaringan-extra, echo=FALSE, warning=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ))
rmarkdown::html_dependency_font_awesome()
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE)

cities <- readr::read_csv(file = here::here("week-05/data/cities.csv"))
```

### In class activity

- This time we will start with an activity.

--

- Download the data file `cities.csv` and the .Rmd file for the in class 
activity and solve the first three points.

---
### Example: average temperature

```{r, echo=FALSE}

miami <- subset(x = cities, subset = city == "miami")
sfn <- subset(x = cities, subset = city == "san_francisco")

par(mai = c(0.1, 0.1, 0.1, 0.1), 
    oma = c(3, 3, 0.5, 0.5))

plot(0,0,ann = FALSE, type = "n", las = 1, bty = "l",
     xlim = c(1, 12), ylim = c(40, 90))

qq_m <- miami |> 
  dplyr::group_by(month) |> 
  dplyr::summarise("q1" = quantile(x = avgtemp, probs = 0.25),
                   "q2" = quantile(x = avgtemp, probs = 0.5),
                   "q3" = quantile(x = avgtemp, probs = 0.75),
                   "max" = max(avgtemp),
                   "min" = min(avgtemp))

qq_s <- sfn |> 
  dplyr::group_by(month) |> 
  dplyr::summarise("q1" = quantile(x = avgtemp, probs = 0.25),
                   "q2" = quantile(x = avgtemp, probs = 0.5),
                   "q3" = quantile(x = avgtemp, probs = 0.75),
                   "max" = max(avgtemp),
                   "min" = min(avgtemp))

rect(xleft = seq(from = 0.7, to = 11.7, by = 1), 
     xright = seq(from = 1.3, 12.3, by = 1),
     ybottom = qq_m$q1, ytop = qq_m$q3, col = "#ff6347aa", 
     border = "tomato")

segments(x0 = seq(from = 0.7, to = 11.7, by = 1),
         x1 = seq(from = 1.3, 12.3, by = 1),
         y0 = qq_m$q2,
         y1 = qq_m$q2, 
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, 12, by = 1),
         y0 = qq_m$q3,
         y1 = ifelse(test = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1) > qq_m$max, 
                     yes = qq_m$max,
                     no = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, to = 12, by = 1),
         y0 = qq_m$q1,
         y1 = ifelse(test = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1) < qq_m$min, 
                     yes = qq_m$min,
                     no = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

rect(xleft = seq(from = 0.5, to = 11.5, by = 1), 
     xright = seq(from = 1.1, 12.1, by = 1),
     ybottom = qq_s$q1, ytop = qq_s$q3, col = "#1e90ffaa", 
     border = "dodgerblue")

segments(x0 = seq(from = 0.5, to = 11.5, by = 1),
         x1 = seq(from = 1.1, 12.1, by = 1),
         y0 = qq_s$q2,
         y1 = qq_s$q2, 
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q3,
         y1 = ifelse(test = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1) > qq_s$max, 
                     yes = qq_s$max,
                     no = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q1,
         y1 = ifelse(test = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1) < qq_s$min, 
                     yes = qq_s$min,
                     no = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

mtext(text = "Month", side = 1, line = 2.3, cex = 1.5)
mtext(text = "Average temperature", side = 2, line = 2.3, cex = 1.5)
legend("bottom", legend = c("Miami", "San Francisco"), pch = 15, 
       bty = "n", col = c("tomato", "dodgerblue"), cex = 1.3)

```

- What can we tell from the plot?

---
### Example: average temperature

- It is clear from the plot that the average daily temperature does not follow 
a straight line.

--

- This actually agrees with our experience, at least in the northern hemisphere 
the months of July, August, and September are usually the warmer of the year.

--

- Then around October and November the temperature starts to drop.

--

- Can we still use a linear model to account for the variability in average 
temperature in each city?

--

- It turns out that we can, however, it is more difficult than just using a 
multiple linear regression.

---
### A general version of the linear model

- Up to this point we have talked about the linear model in a simplified way. 

--

- In each of our models we have a set of covariates that we want to use in 
order to account for the expected value of some dependent variable.

--

- However, we haven't asked ourselves what makes a "linear model" linear.

--

- For example, is the following model a linear model?
$$\mathrm{E}\left(Y_i\mid X_{i1}\right) = \beta_0 + \beta_1e^{X_{i1}}$$

--

- Well, it turns out that this is a linear model. But why?

---
### A general version of the linear model

- A model is considered linear if the parameters and covariates are combined 
using only addition and multiplication.

--

- In our previous example, even though we have a transformation of our 
covariate $e^{X_{i1}}$, it is still being multiplied by the parameter $\beta_1$ 
and that quantity is added to another parameter $\beta_0$.

--

- Notice that we have not said that the covariates have to be used as they
have been measured in an experiment.

--

- Only that we need to combine whatever form of the covariate with the 
parameters using only addition and multiplication.

---
### A general version of the linear model

- In general, a linear model can be expressed as:
$$Y_i = \beta_0 + \beta_1 f_1(X_{i1}) + \cdots +  \beta_k f_k(X_{i1}) + \beta_{k+1} f_{k+1}(X_{i2}) + \cdots + \beta_K f_{K}(X_{ip}) + \epsilon_i$$

--

- What does this mean?

--

- Well it means that we can use any transformation of the covariates that we 
have in the sample $X_{i1},\dots, X_{ip}$.

--

- This brings another level of flexibility to a linear models and at the same 
time increases the number of decisions that we need to make.

--

- For example, should we include the month to our average temperature model or
should we add some transformation of the month?

---
### Finding transformations

- Using covariate transformations will modify the predictions that the model 
can make from a straight line into a curve.

--

- Finding which transformation to use is not easy and normally we would prefer 
to justify their addition using theory.

--

- However, if we do not have the support of a theory then we can look at the 
residuals and try to find a way to deal with the lack of linearity in the data.

--

- Let's look at the example using the two cities data.

--

- First let's fit a model that includes only the city as a predictor and look 
at the residuals.

--

```{r}
lm_city <- lm(formula = avgtemp ~ city, data = cities) 
```

---
### Example: average temperature

- According to our model, the estimated expected average temperature for the 
city of Miami in 2017 was `r round(x = lm_city$coef[1], digits = 2)` degrees 
Fahrenheit.

- The estimated expected average temperature on the city of San Francisco was 
`r round(x = lm_city$coef[1] + lm_city$coef[2], digits = 2)` degrees 
Fahrenheit, which is `r round(x = -lm_city$coef[2], digits = 2)` 
degrees below the estimated value for Miami.

```{r, echo=FALSE, out.width="37%"}
par(mai = c(0.1, 0.1, 0.1, 0.1), 
    oma = c(3, 3, 0.5, 0.5))

plot(0,0,ann = FALSE, type = "n", las = 1, bty = "l",
     xlim = c(1, 12), ylim = c(40, 90))

rect(xleft = seq(from = 0.7, to = 11.7, by = 1), 
     xright = seq(from = 1.3, 12.3, by = 1),
     ybottom = qq_m$q1, ytop = qq_m$q3, col = "#ff6347aa", 
     border = "tomato")

segments(x0 = seq(from = 0.7, to = 11.7, by = 1),
         x1 = seq(from = 1.3, 12.3, by = 1),
         y0 = qq_m$q2,
         y1 = qq_m$q2, 
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, 12, by = 1),
         y0 = qq_m$q3,
         y1 = ifelse(test = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1) > qq_m$max, 
                     yes = qq_m$max,
                     no = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, to = 12, by = 1),
         y0 = qq_m$q1,
         y1 = ifelse(test = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1) < qq_m$min, 
                     yes = qq_m$min,
                     no = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

rect(xleft = seq(from = 0.5, to = 11.5, by = 1), 
     xright = seq(from = 1.1, 12.1, by = 1),
     ybottom = qq_s$q1, ytop = qq_s$q3, col = "#1e90ffaa", 
     border = "dodgerblue")

segments(x0 = seq(from = 0.5, to = 11.5, by = 1),
         x1 = seq(from = 1.1, 12.1, by = 1),
         y0 = qq_s$q2,
         y1 = qq_s$q2, 
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q3,
         y1 = ifelse(test = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1) > qq_s$max, 
                     yes = qq_s$max,
                     no = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q1,
         y1 = ifelse(test = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1) < qq_s$min, 
                     yes = qq_s$min,
                     no = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

abline(h = c(lm_city$coef[1], lm_city$coef[1] + lm_city$coef[2]), lwd = 4,
       col = c("red", "blue"), lty = c(1,2))

mtext(text = "Month", side = 1, line = 2.3, cex = 1.5)
mtext(text = "Average temperature", side = 2, line = 2.3, cex = 1.5)
legend("bottom", legend = c("Miami", "San Francisco"), pch = 15, 
       bty = "n", col = c("tomato", "dodgerblue"), cex = 1.3)
```

---
### Example: average temperature

- Now let's look at the residuals as a function of the month ans see what we 
can find.

```{r, echo=FALSE, out.width="50%"}
par(mai = c(0.1, 0.1, 0.1, 0.1), 
    oma = c(3, 3, 0.5, 0.5))

plot(0,0,ann = FALSE, type = "n", las = 1, bty = "l",
     xlim = c(1, 12), 
     ylim = c(min(lm_city$residuals), max(lm_city$residuals)))

points(x = jitter(cities$month, amount = 0.3), y = lm_city$residuals,
       pch = 21, bg = "slategray", cex = 1.2)
abline(h = 0, lwd = 3, lty = 2)

mtext(text = "Month", side = 1, line = 2.3, cex = 1.5)
mtext(text = "Residuals", side = 2, line = 2.3, cex = 1.5)
```

---
### Example: average temperature

- From the residuals (as well as the data plot) we can see the non-linearity in 
the data.

--

- However, we can also see that that model is over estimating our observations 
most of the time. 

--

- One thing we can do to improve the model is just add the month covariate to 
the regression without any transformation.

--

- Adding the month will help us to reduce the bias of the model ar least on the
first couple of months.

--

- How will this happen? well remember that the intercept and the slope are 
correlated, so adding a slope will change how the intercept is estimated.

```{r}
lm_city_month <- lm(formula = avgtemp ~ city + month, data = cities)
```

---
### Example: average temperature

- Now we can look at the expected value predicted by the model against the data

```{r, echo=FALSE, out.width="50%"}
par(mai = c(0.1, 0.1, 0.1, 0.1), 
    oma = c(3, 3, 0.5, 0.5))

plot(0,0,ann = FALSE, type = "n", las = 1, bty = "l",
     xlim = c(1, 12), ylim = c(40, 90))

rect(xleft = seq(from = 0.7, to = 11.7, by = 1), 
     xright = seq(from = 1.3, 12.3, by = 1),
     ybottom = qq_m$q1, ytop = qq_m$q3, col = "#ff6347aa", 
     border = "tomato")

segments(x0 = seq(from = 0.7, to = 11.7, by = 1),
         x1 = seq(from = 1.3, 12.3, by = 1),
         y0 = qq_m$q2,
         y1 = qq_m$q2, 
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, 12, by = 1),
         y0 = qq_m$q3,
         y1 = ifelse(test = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1) > qq_m$max, 
                     yes = qq_m$max,
                     no = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, to = 12, by = 1),
         y0 = qq_m$q1,
         y1 = ifelse(test = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1) < qq_m$min, 
                     yes = qq_m$min,
                     no = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

rect(xleft = seq(from = 0.5, to = 11.5, by = 1), 
     xright = seq(from = 1.1, 12.1, by = 1),
     ybottom = qq_s$q1, ytop = qq_s$q3, col = "#1e90ffaa", 
     border = "dodgerblue")

segments(x0 = seq(from = 0.5, to = 11.5, by = 1),
         x1 = seq(from = 1.1, 12.1, by = 1),
         y0 = qq_s$q2,
         y1 = qq_s$q2, 
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q3,
         y1 = ifelse(test = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1) > qq_s$max, 
                     yes = qq_s$max,
                     no = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q1,
         y1 = ifelse(test = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1) < qq_s$min, 
                     yes = qq_s$min,
                     no = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

abline(a = lm_city_month$coef[1], b = lm_city_month$coef[3], lwd = 4,
       col = "red", lty = 1)

abline(a = lm_city_month$coef[1] + lm_city_month$coef[2],
       b = lm_city_month$coef[3], 
       lwd = 4, col = "blue", lty = 2)

mtext(text = "Month", side = 1, line = 2.3, cex = 1.5)
mtext(text = "Average temperature", side = 2, line = 2.3, cex = 1.5)
legend("bottom", legend = c("Miami", "San Francisco"), pch = 15, 
       bty = "n", col = c("tomato", "dodgerblue"), cex = 1.3)
```

---
### Example: average temperature

- Once again we can look at the residuals to see what has changed.

```{r, echo=FALSE, out.width="50%"}
par(mai = c(0.1, 0.1, 0.1, 0.1), 
    oma = c(3, 3, 0.5, 0.5))

plot(0,0,ann = FALSE, type = "n", las = 1, bty = "l",
     xlim = c(1, 12), 
     ylim = c(min(lm_city_month$residuals), max(lm_city_month$residuals)))

points(x = jitter(cities$month, amount = 0.3), y = lm_city_month$residuals,
       pch = 21, bg = "slategray", cex = 1.2)
abline(h = 0, lwd = 3, lty = 2)

mtext(text = "Month", side = 1, line = 2.3, cex = 1.5)
mtext(text = "Residuals", side = 2, line = 2.3, cex = 1.5)
```

---
### Example: average temperature

- The residuals show a lower bias for the first few months, however, we can 
still see that there is a non-linear trend and that the temperature for the 
last few months is being overestimated.

--

- Now we can start thinking about how to transform the values of the month in 
order to get the effect we want.

--

- First let's look at the trends in the residuals to know which type of 
function we would want to use.

--

- From the plot of the residuals we can see that they are lower at the beginning
and the end of the year. 

--

- At the same time, the highest values are around month 6 or 7.

---
### Cosine function

- If we assume that those values are symmetric what we want is a function that 
assigns small or negative values to the first and last months and assigns
positive values near to month 7.

--

- One function that can do something like this is the cosine function.

--

- A cosign function assigns a value of 1 to values close to 0 or 2$\pi$, while
it assigns values near -1 to values around $\pi < x < 3/2 \pi$.

--

- The first thing we could do is to change our month values so that they start 
at 0 and end at $2\pi$ this way we make sure that both January and December 
will have the largest values once we use the cosine function.

--

- We can transform our data using the following function:
$$- \operatorname{cos}\left( 2\pi \frac{month_i - 1}{11}\right)$$

---
### Cosine function

- What does this transformation look like?

```{r, echo=FALSE}
curve(expr = -cos(2*pi*((x-1)/11)), from = 1, to = 12, axes = FALSE, 
      ann = FALSE, lwd = 3)
box(bty = "l")
axis(1)
axis(2)
mtext(text = "Month", side = 1, line = 2.3)
```

---
### Cosine function

- As we can see, this function assigns the lowest values to months near 1 and 12

--

- At the same time it assigns the largest values to months near 6.

--

- We can use this transformation in our model, but first we need to add it to 
the data

```{r}
cities <- cities |> 
  dplyr::mutate("cos_month" = -cos(2 * pi * ((month - 1) / 11)))
```

- This adds a new column to our data called `cos_month` that has the value of 
the transformation for each row in our data frame.

--

- Now that we have our new variable we can use it in a new linear model

```{r}
lm_transform <- lm(formula = avgtemp ~ city + month + cos_month, 
                   data = cities)
```

---
### Example: average temperature

- As we have done before, we can look at the predicted values from the linear 
model

```{r, echo=FALSE, out.width="50%"}
par(mai = c(0.1, 0.1, 0.1, 0.1), 
    oma = c(3, 3, 0.5, 0.5))

plot(0,0,ann = FALSE, type = "n", las = 1, bty = "l",
     xlim = c(1, 12), ylim = c(40, 90))

rect(xleft = seq(from = 0.7, to = 11.7, by = 1), 
     xright = seq(from = 1.3, 12.3, by = 1),
     ybottom = qq_m$q1, ytop = qq_m$q3, col = "#ff6347aa", 
     border = "tomato")

segments(x0 = seq(from = 0.7, to = 11.7, by = 1),
         x1 = seq(from = 1.3, 12.3, by = 1),
         y0 = qq_m$q2,
         y1 = qq_m$q2, 
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, 12, by = 1),
         y0 = qq_m$q3,
         y1 = ifelse(test = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1) > qq_m$max, 
                     yes = qq_m$max,
                     no = qq_m$q3 + (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

segments(x0 = seq(from = 1, to = 12, by = 1),
         x1 = seq(from = 1, to = 12, by = 1),
         y0 = qq_m$q1,
         y1 = ifelse(test = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1) < qq_m$min, 
                     yes = qq_m$min,
                     no = qq_m$q1 - (1.5 * qq_m$q3 - qq_m$q1)),
         col = "tomato", 
         lwd = 2)

rect(xleft = seq(from = 0.5, to = 11.5, by = 1), 
     xright = seq(from = 1.1, 12.1, by = 1),
     ybottom = qq_s$q1, ytop = qq_s$q3, col = "#1e90ffaa", 
     border = "dodgerblue")

segments(x0 = seq(from = 0.5, to = 11.5, by = 1),
         x1 = seq(from = 1.1, 12.1, by = 1),
         y0 = qq_s$q2,
         y1 = qq_s$q2, 
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q3,
         y1 = ifelse(test = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1) > qq_s$max, 
                     yes = qq_s$max,
                     no = qq_s$q3 + (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

segments(x0 = seq(from = 0.8, to = 11.8, by = 1),
         x1 = seq(from = 0.8, to = 11.8, by = 1),
         y0 = qq_s$q1,
         y1 = ifelse(test = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1) < qq_s$min, 
                     yes = qq_s$min,
                     no = qq_s$q1 - (1.5 * qq_s$q3 - qq_s$q1)),
         col = "dodgerblue", 
         lwd = 2)

curve(expr = lm_transform$coef[1] + lm_transform$coef[3] * x +
        lm_transform$coef[4] * -cos(2 * pi * ((x - 0.5) / 11)),
      lwd = 4, col = "red", lty = 1, add = TRUE)

curve(expr = lm_transform$coef[1] + lm_transform$coef[2] +
        lm_transform$coef[3] * x +
        lm_transform$coef[4] * -cos(2 * pi * ((x - 0.5) / 11)),
      lwd = 4, col = "blue", lty = 2, add = TRUE)

mtext(text = "Month", side = 1, line = 2.3, cex = 1.5)
mtext(text = "Average temperature", side = 2, line = 2.3, cex = 1.5)
legend("bottom", legend = c("Miami", "San Francisco"), pch = 15, 
       bty = "n", col = c("tomato", "dodgerblue"), cex = 1.3)
```

---
### Example: average temperature

- Adding the transformed value of the months actually improves the model

--

- But is the improvement enough?

--

- We can use the BIC to see if the model that includes the transformed month 
is better than one that does not include it

--

- The BIC of the model that does not include the transformation was 
`r round(x = BIC(lm_city_month), digits = 2)`, while the BIC for the model that 
does include the transformation was `r round(x = BIC(lm_transform), digits = 2)`

--

- Given that the BIC for the model that includes the transformation is lower 
we can say that his model improves on the one that only has the city and the 
month as covariates.

--

- However, there is still a problem. If we look at the residuals we can see
that there is still bias in our estimation of the expected average temperature 
by month.

---
### Example: average temperature

```{r, echo=FALSE, out.width="30%"}
par(mai = c(0.1, 0.1, 0.1, 0.1), 
    oma = c(3, 3, 0.5, 0.5))

plot(0,0,ann = FALSE, type = "n", las = 1, bty = "l",
     xlim = c(1, 12), 
     ylim = c(min(lm_transform$residuals), max(lm_transform$residuals)))

points(x = jitter(cities$month, amount = 0.3), y = lm_transform$residuals,
       pch = 21, bg = "slategray", cex = 1.2)
abline(h = 0, lwd = 3, lty = 2)

mtext(text = "Month", side = 1, line = 2.3, cex = 1.5)
mtext(text = "Residuals", side = 2, line = 2.3, cex = 1.5)
```

- With the information we have there is not a lot we can do to solve this 
problem, we could add other transformations or covariates.

--

- Choosing transformations for covariates can be useful in a linear regression, 
however, the problem of which transformation to use is difficult to solve and 
most of the time we will try to avoid it.



