---
title: "Non-Constant Variance"
subtitle: "Lecture 10"
author: 
  - "Manuel Villarreal"
date: '09/30/24'
output:
  xaringan::moon_reader:
    self_contained: true
    css: ['xaringan-themer.css', 'set-colors.css']
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.width = 7,
  fig.height = 5.8,
  fig.align = 'center',
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, echo=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#005f86",
  secondary_color = "#f8971f",
  inverse_header_color = "#FFFFFF"
)
```

```{r xaringan-extra, echo=FALSE, warning=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ))
rmarkdown::html_dependency_font_awesome()
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE)
```
### Assumptions of the linear model

- Remember the assumptions of the linear model:

--

  1. Errors have an expectation of 0 $\mathrm{E}(\epsilon) = 0$.
  
--

  1. Errors are identically distributed 
  $\epsilon_i \sim \left(\mathrm{E}(\epsilon), \mathrm{Var}(\epsilon)\right)$.
  
--

  1. Errors are independent $\epsilon_i \perp \epsilon_j$ for all $i \neq j$
  in $1, \dots, n$.

--

  1. Errors have constant variance $\mathrm{Var}(\epsilon_i) = \sigma^2$.
  
--

- When all of these assumptions are met the OLS estimators for $\beta$:
$$\hat\beta \dot\sim \mathrm{Normal}(\beta, \mathrm{Var}(\hat\beta))$$

---
### Variance of the OLS

- From the previous lecture we know that the when all assumptions are met 
including the constant variance, then the variance of our OLS estimators is:
$$\mathrm{Var}(\hat\beta) = \sigma^2\left(X^TX\right)^{-1}$$

--

- This is a direct result of the constant variance assumption which can be 
expressed as:
$$\mathrm{Var}(\epsilon) = \sigma^2\boldsymbol{I}_{n \times n}$$

--

$$\mathrm{Var}(\epsilon) = \begin{pmatrix}
\sigma^2 & 0        & \cdots & 0      \\
0        & \sigma^2 & \cdots & 0      \\
\vdots   & \vdots   & \ddots & \vdots \\
0        & 0        & \cdots & \sigma^2
\end{pmatrix}$$

---
### Consequences of non-constant variance

- What happens when the constant variance assumption is not met?

--

- We can start by formalizing the problem, in this case it would mean that 
$\mathrm{Var}(\epsilon) \neq \sigma^2\boldsymbol{I}$. Therefore, we need to 
start by specifying the variance of the errors.

--

- Let's say for now that we still meet the assumption of independence, that is 
$\epsilon_i \perp \epsilon_j$. When the errors are independent then we know that
their covariance will be equal to $0$.

--

- That means that we can express the variance in matrix form as:
$$\mathrm{Var}(\epsilon) = \begin{pmatrix}
\sigma^2_1 & 0        & \cdots & 0      \\
0        & \sigma^2_2 & \cdots & 0      \\
\vdots   & \vdots   & \ddots & \vdots \\
0        & 0        & \cdots & \sigma^2_n
\end{pmatrix} = \Sigma_{n\times n}$$

---
### Consequences of non-constant variance

- Here $\Sigma_{n\times n}$ represents the variance covariance matrix of our 
errors.

--

- If the other assumptions are true, then that means that for a linear model of 
the form $Y = X\beta + \epsilon$ we have that:

--

  1. $\mathrm{E}(Y) = \mathrm{E}(X\beta + \epsilon) = X\beta + \mathrm{E}(\epsilon) =  X\beta$

--

 1. $\mathrm{Var}(Y) = \mathrm{Var}(X\beta + \epsilon) = \mathrm{Var}(\epsilon) =  \Sigma$
 
--

- In other words, even though we have lost the assumption of constant variance, 
the mean of our observations should still be the linear function $X\beta$.

---
### Consequences of non-constant variance

- Now, what happens with our OLS estimators $\hat\beta$, well they are still
unbiased, which means that their expected value is still equal to the population
level parameters $\beta$:

--

- $\mathrm{E}\left(\hat\beta\right) = \mathrm{E}\left(\left(X^TX\right)^{-1}X^TY\right) = \left(X^TX\right)^{-1}X^T\mathrm{E}\left(Y\right) = \left(X^TX\right)^{-1}X^TX\beta = \beta$

--

- However, the problem show's up when we look at the variance of our estimators:
$$\mathrm{Var}\left(\hat\beta\right) = \mathrm{Var}\left(\left(X^TX\right)^{-1}X^TY\right)\\
\mathrm{Var}\left(\hat\beta\right) = \left(X^TX\right)^{-1}X^T\mathrm{Var}\left(Y\right)X\left(X^TX\right)^{-1}\\
\mathrm{Var}\left(\hat\beta\right) = \left(X^TX\right)^{-1}X^T\Sigma X\left(X^TX\right)^{-1}$$

---
### Consequences of non-constant variance

- The problem here is that now the variance of our errors $\Sigma$ is there and
we can't simplify the variance any further.

--

- In practice, this means that we can't use the Sum of Squared Errors 
$\hat\sigma^2$ to estimate the variance of our estimators.

--

- This means that we can't get the standard error of our estimators with the 
formula that we have used before.

--

- We now have to take extra steps to approximate the variance of our estimators
if we want to do inferences with them.

--

- But, what would happen if we just ignore that this is true and instead keep 
using the same standard error (which are the default in R)?

--

- Well let's look at the consequences using a simulation example.

---
### Simulation

- Open the file `in-class-11.Rmd` and follow the instructions.

--

- First let's start by generating the mean of our observations using 5 different
values of X.

--

- Generate the mean of a random variable $\mu$ using the linear function for
5 different values of x that go from 1 to 5 each repeated 6 times. You can 
choose any values for the parameters $\beta_0$ and $\beta_1$ 
(question 1 `in-class-11.Rmd`).

--

```{r}
# set values for the parameters beta_0 and beta_1
beta <- c(1, 3)

# generate a design matrix with for 30 observations in which the first column
# has the value of 1 and the second column has the value of our predictor
design <- cbind(rep(1, times = 5 * 6), 
                rep(seq(from = 1, to = 5), each = 6))

# generate a vector with the mean for each observation
mu <- design %*% beta
```

---
### Simulation

- Notice that the mean we generated is the TRUE mean, and not the estimated one. 
This is because in a simulation we know the true values of the parameters, and 
therefore the true mean.

--

- Now that we have a mean vector we have to generate a vector of variances to
simulate the non-constant variance problem.

--

- Generate a vector of variances named `variance_obs` using the following format
$\mathrm{Var}(\epsilon_i\mid x) = 1.5 + \frac{x - 1}$ 
(question 2 `in-class-11.Rmd`).

--

```{r}
# generate a vector with as many variances as mean values generated in 1.
variance_obs <- 1.5 + (design[, 2] - 1)
```



---
### Simulation

- Again, because this is a simulation we can set the values of x from the start 
and then treat it as an unknown quantity once we have the data.

--

- Now that we have our vector of means $\mu = \boldsymbol{X}\beta$ and our 
vector of variances $\sigma^2_1, \dots, \sigma^2_n$ we can generate data from a
single experiment.

--

- Generate a vector of 30 observations $Y$ with the following properties:
$$y_i \sim \mathrm{Normal}(\beta_0 + \beta_1x_{i1}, \sigma^2_i)$$ (*Note:* you 
can do this using the function `rnorm` and the variables from questions 1 and 2).

--

```{r}
# generate data from a single experiment with 30 observations with mean 
# mu and variance sigma^2_i
observations <- rnorm(n = dim(design)[1], mean = mu, sd = sqrt(variance_obs))
```

---
### Simulation

- Make a plot of the data you just generated and add the regression line 
(question 4, `in-class-11.Rmd`).

--

```{r, echo=FALSE}
plot(x = jitter(x = design[, 2], amount = 0.2), 
     y = observations, las = 1, ann = FALSE, pch = 21, col = "#FFFFFF", 
     bg = "#f8971f", cex = 2, bty = "l")
abline(a = beta[1], b = beta[2], col = "#005f86", lwd = 3, lty = 2)
mtext(text = "Predictor", side = 1, line = 2.4, cex = 1.6)
mtext(text = "Observation", side = 2, line = 3, cex = 1.3)
```

---
### Simulation

- Fit a linear model to your observations and compare the estimated values 
$\hat\beta_0$ and $\hat\beta_1$ with the real values used in your simulation
(question 5, `in-class-11.Rmd`).

--

```{r}
# fitting a linear model using the predictor in our design matrix
lm_sim <- lm(formula = observations ~ design[, 2])
```

--

- The estimated value of $\hat\beta_0$ was
`r round(x = lm_sim$coef[1], digits = 2)` which is approximately 
`r round(x = abs(beta[1] - lm_sim$coef[1]), digits = 3)` away from the true 
value.

--

- The estimated value of $\hat\beta_1$ was
`r round(x = lm_sim$coef[2], digits = 2)` which is approximately 
`r round(x = abs(beta[2] - lm_sim$coef[2]), digits = 3)` away from the true 
value.

---
### Simulation

- Plot the residuals of the model as a function of the predictor and describe 
the plot, including any trends that you can see (question 6, `in-class-11.Rmd`).

--

```{r, echo=FALSE}
plot(x = jitter(x = design[, 2], amount = 0.2),
     y = lm_sim$residuals, las = 1, ann = FALSE,
     pch = 21, col = "#FFFFFF", bg = "#f8971f",
     cex = 2, bty = "l")
abline(h = 0, col = "#005f86", lwd = 3, lty = 2)
mtext(text = "Predictor", side = 1, line = 2.4, cex = 1.6)
mtext(text = "Residuals", side = 2, line = 3, cex = 1.3)
```

---
### Simulation

- Now let's simulate 10,000 experiments, in this section we will write the 
simulation together so that I can show you what to save and how to make the 
simulation run faster (question 7, `in-class-11.Rmd`).

```{r, echo=FALSE}
inv_design <- solve(t(design) %*% design)

n_simulations <- 10000
sample_size <- dim(design)[1]
n_parameters <- dim(design)[2]

hat_beta_0 <- c()
hat_beta_1 <- c()
hat_sigma <- c()

for (i in 1:n_simulations) {
  y <- rnorm(n = dim(design)[1], mean = mu, sd = sqrt(variance_obs))
  hat_beta <- inv_design %*% t(design) %*% y
  hat_sigma <- append(
    x = hat_sigma, 
    values = sum((y - design %*% hat_beta)^2) / (sample_size - n_parameters))
  hat_beta_0 <- append(x = hat_beta_0, values = hat_beta[1])
  hat_beta_1 <- append(x = hat_beta_1, values = hat_beta[2])
}
```

---
### Simulation

- Make a histogram of the estimated values of $\hat\beta_0$, use the argument
`freq = FALSE` to have the y axis be a density and not a count 
(question 8, `in-class-11.Rmd`).

--

```{r, echo=FALSE}
hist(x = hat_beta_0, freq = FALSE, breaks = 100, las = 1, ann = FALSE,
     border = "#ffffff", col = "#f8971f")
mtext(text = expression(beta[0]), side = 1, line = 2.4, cex = 1.6)
abline(v = beta[1], lwd = 3, lty = 2, col = "#005f86")
```

---
### Simulation

- Make a histogram of the estimated values of $\hat\beta_1$, use the argument
`freq = FALSE` to have the y axis be a density and not a count 
(question 9, `in-class-11.Rmd`).

--

```{r, echo=FALSE}
hist(x = hat_beta_1, freq = FALSE, breaks = 100, las = 1, ann = FALSE,
     border = "#ffffff", col = "#f8971f")
mtext(text = expression(beta[1]), side = 1, line = 2.4, cex = 1.6)
abline(v = beta[2], lwd = 3, lty = 2, col = "#005f86")
```

---
### Non-constant Variance

- Up to this point everything still looks good, the distributions of the 
simulated OLS estimators even have a normal distribution.

--

- Now let's calculate the confidence interval for each of the simulations.

--

- Calculate the $95\%$ confidence interval for each $\hat\beta_0$ obtained
in the simulation, the result should be a matrix with 10,000 rows (one for 
each simulation) and two columns, one for the lower and upper bound 
respectively (question 10, `in-class-11.Rmd`).

--

```{r}
alpha <- 0.05
beta_0_ci <- cbind(
  hat_beta_0 - qt(p = 1 - alpha / 2, df = sample_size - n_parameters) * 
    sqrt(hat_sigma * inv_design[1, 1]),
  hat_beta_0 + qt(p = 1 - alpha / 2, df = sample_size - n_parameters) * 
    sqrt(hat_sigma * inv_design[1, 1]))
```

---
### Non-constant Variance

- Now that we have the $95\%$ confidence interval for $\beta_0$ we can look at
how many of those intervals contain the true value of the parameter.

--

- Calculate the proportion of times that the true value of the parameter 
$\beta_0$ is contained between the lower and upper bound you calculated. Is this
proportion greater, lower or equal to the theoretical coverage probability 
$(0.95)$. (question 11, `in-class-11.Rmd`).

--

- We can estimate the observed coverage probability by taking the mean of a
logical operation. Remember that `TRUE` is treated as 1 in R.

```{r}
coverage_b0 <- mean(x = beta_0_ci[, 1] <= beta[1] & beta[1] <= beta_0_ci[, 2])
```

- The proportion of intervals that contained the true value of the parameter 
$\beta_0$ was `r round(x = coverage_b0, digits = 3)`, this is higher than the 
theoretical coverage probability we derived from the theory.

---
### Non-constant Variance

- Calculate the $95\%$ confidence interval for each $\hat\beta_1$ obtained
in the simulation, the result should be a matrix with 10,000 rows (one for 
each simulation) and two columns, one for the lower and upper bound 
respectively. Like we did with $\beta_0$ estimate the proportion of times that 
the true value of the parameter $\beta_1$ is contained between the two bounds. 
Is the proportion greater, lower or equal to the theoretical coverage 
probability $(0.95)$ (question 12, `in-class-11.Rmd`).

--

```{r}
alpha <- 0.05
beta_1_ci <- cbind(
  hat_beta_1 - qt(p = 1 - alpha / 2, df = sample_size - n_parameters) * 
    sqrt(hat_sigma * inv_design[2, 2]),
  hat_beta_1 + qt(p = 1 - alpha / 2, df = sample_size - n_parameters) * 
    sqrt(hat_sigma * inv_design[2, 2]))

coverage_b1 <- mean(x = beta_1_ci[, 1] <= beta[2] & beta[2] <= beta_1_ci[, 2])
```

---
### Non-constant Variance

- The proportion of times that the interval for $\beta_1$ contained the true 
value of the parameter was `r round(x = coverage_b1, digits = 3)`, which is 
close to the theoretical coverage probability we derived from the theory.

--

- In this case we have an example where the coverage probability for the 
parameter $\beta_0$ is the one that suffers the most. However, this does not
mean that the unequal variance problem will only affect the inference for 
the intercept. It can affect all the parameters in a model.

--

- The key issued here is that the variance as we have derive it (assuming that
observations have a constant variance), is overestimating the variability of the 
intercept which results in an increased confidence interval and therefore a 
higher coverage probability than expected.

---
### Solutions to the Heteroskedasticity Problem

- Some solutions have been introduced to deal with the problem of unequal 
variances. One such solution is the use of `Robust Variance Estimators`.

--

- A robust variance estimator is a function that attempts to correct the bias in
our estimation of the variance of the OLS estimators.

--

- There are multiple estimators (functions) in these family of estimators, 
however, one that is often used is the 
`Heteroskedasticity Corrected Variance Estimator`.

--

- The key intuition of this approach is to use the squared value of the 
residual for each observation (the difference between the observation and the 
mean) as an estimate of $\sigma^2_i$ and then correct this value by taking 
into account how much influence a particular observation has on the predictions
of the model.

---
### Measuring the impact of each observation on predicted values

- In a linear model we can measure the effect that each observation has on the 
predicted mean $\hat\mu$. 

--

- This is measured by a matrix known as the `Hat matrix` which is defined as:
$$H = \boldsymbol{X}\left(\boldsymbol{X}^T\boldsymbol{X}\right)^{-1}\boldsymbol{X}^T$$

--

- The reason why is called the hat matrix is because when we multiply this
matrix by the observations $\boldsymbol{Y}$ the result is the `predicted` mean 
of each observation. Which we typically denote as $\hat{\mu}$.

---
### Measuring the impact of each observation on predicted values

- It turns out that the elements in the main diagonal of the hat matrix $H$ are 
an indicator of how much influence each observation has on the expected value of 
y.

--

- These diagonal elements can only take values between $0$ and $1$, and the 
closer they are to $1$ the more influential they are.

--

- Create a new vector named `h` that includes the diagonal elements of the 
hat matrix for your simulations. **Notice** that the hat matrix does not change 
across simulations of the experiment (question 13, `in-class-11.Rmd`).

--

```{r}
h <- diag(x = design %*% inv_design %*% t(design))
```

---
### Heteroskedasticity Corrected Variance Estimator

- Now that we have our measure of the influence of each observation on the 
regression line we will use it to estimate a new variance matrix.

--

- Remember that we said that we could write the variance of the observations, 
or the variance of the errors as:

$$\mathrm{Var}(\epsilon) = \begin{pmatrix}
\sigma^2_1 & 0        & \cdots & 0      \\
0        & \sigma^2_2 & \cdots & 0      \\
\vdots   & \vdots   & \ddots & \vdots \\
0        & 0        & \cdots & \sigma^2_n
\end{pmatrix}$$

--

- We are going to estimate each $\sigma_i^2$ using:
$$\hat\sigma_i^2 = (y_i - \hat\mu_i)^2$$

---
### Heteroskedasticity Corrected Variance Estimator

- Now that we have both elements we will generate the estimated variance for 
the observations using:

$$\widehat{\mathrm{Var}(\epsilon)} = \begin{pmatrix}
\frac{\hat{\sigma^2_1}}{(1-h_{11})^2} & 0        & \cdots & 0      \\
0        & \frac{\hat{\sigma^2_2}}{(1-h_{22})^2} & \cdots & 0      \\
\vdots   & \vdots   & \ddots & \vdots \\
0        & 0        & \cdots & \frac{\hat{\sigma^2_n}}{(1-h_{nn})^2}
\end{pmatrix}$$

--

- Where $h_{ii}$ represents the $i-th$ element in the main diagonal of H.

---
### Heteroskedasticity Corrected Variance Estimator

- Let's see what happens to a confidence interval when we use the new 
variance estimator $\widehat{\mathrm{Var}(\epsilon)}$. Calculate the matrix 
$\widehat{\mathrm{Var}(\epsilon)}$ and save it with the name `hat_residuals` 
(`hint:` if you use the function `diag` and use a vector instead of a matrix as 
the argument the result with be a diagonal matrix that includes the elements of 
the vector in the main diagonal; question 14, `in-class-11.Rmd`).

--

- For our original simulation of the data we can get the new variance matrix 
with:

```{r}
hat_residuals <- diag(x = as.vector((observations - mu)^2 * (1 / (1 - h)^2)))
```

---
### Heteroskedasticity Corrected Variance Estimator

- Our new matrix $\widehat{\mathrm{Var}(\epsilon)} = \hat\Sigma$ is known as the Type 3
Heteroskedasticity Corrected Variance Estimator, or HC3 for short.

--

- Now we can calculate a corrected version of the variance of our OLS 
estimators: 
$$\widehat{\mathrm{Var}(\hat\beta)} = (\boldsymbol{X}^T\boldsymbol{X})^{-1}\boldsymbol{X}^T\hat\Sigma\boldsymbol{X}(\boldsymbol{X}^T\boldsymbol{X})^{-1}$$

--

- Calculate the corrected variance of the OLS estimators 
(question 15, `in-class-11.Rmd`).

--

```{r}
corrected_var_beta <- 
  inv_design %*% t(design) %*% 
  hat_residuals %*%
  design %*% inv_design
```

---
### Heteroskedasticity Corrected Variance Estimator

- We can use the new estimated variance of our parameters to estimate confidence 
intervals and p-values.

--

- Estimate the $95\%$ confidence interval for the parameters $\beta_0$ and 
$\beta_1$ using the variance matrix that you found in 15 and compare them with 
the ones found using the output of the `summary` function of an `lm` model.

--

```{r}
alpha <- 0.05
summary_sim <- summary(lm_sim)

ci_beta0 <- lm_sim$coef[1] + c(-1, 1) *
  qt(p = 1 - alpha / 2, df = sample_size - n_parameters) *
  sqrt(corrected_var_beta[1,1])

ci_beta0_lm <- lm_sim$coef[1] + c(-1, 1) *
  qt(p = 1 - alpha / 2, df = sample_size - n_parameters) *
  summary_sim$coefficients['(Intercept)', 'Std. Error']
```

---
### Heteroskedasticity Corrected Variance Estimator

- For $\beta_1$ we do:

```{r}
ci_beta1 <- lm_sim$coef[2] + c(-1, 1) *
  qt(p = 1 - alpha / 2, df = sample_size - n_parameters) *
  sqrt(corrected_var_beta[2,2])

ci_beta1_lm <- lm_sim$coef[2] + c(-1, 1) *
  qt(p = 1 - alpha / 2, df = sample_size - n_parameters) *
  summary_sim$coefficients['design[, 2]', 'Std. Error']
```

--

- To compare the values we can make a two plots, each containing a single 
parameter.

```{r, echo=FALSE}
layout(t(c(1,2)))
plot(x = rep(x = 0.25, times = 2), y = ci_beta0, lwd = 4, 
     col = "dodgerblue", ylim = c(-1, 6), type = "l", las = 1, 
     ann = FALSE, axes = FALSE, xlim = c(0,1))
lines(x =  rep(x = 0.75, times = 2), y = ci_beta0_lm, lwd = 4, 
      col = "tomato")
box(bty = "l")
axis(2, las = 2)
mtext(text = expression(beta[0]), side = 1, line = 2.4, cex = 1.6)
legend("topright", bty = "n", legend = c("Corrected", "Constant"),
      col = c("dodgerblue", "tomato"), lwd = 2)

plot(x = rep(x = 0.25, times = 2), y = ci_beta1, lwd = 4, 
     col = "dodgerblue", ylim = c(-1, 6), type = "l", las = 1, 
     ann = FALSE, axes = FALSE, xlim = c(0,1))
lines(x =  rep(x = 0.75, times = 2), y = ci_beta1_lm, lwd = 4, 
      col = "tomato")
box(bty = "l")
axis(2, las = 2)
mtext(text = expression(beta[1]), side = 1, line = 2.4, cex = 1.6)
legend("topright", bty = "n", legend = c("Corrected", "Constant"),
      col = c("dodgerblue", "tomato"), lwd = 2)
```

