---
title: "Simple linear regression: assumptions"
subtitle: "Lecture 2"  
author: 
  - "Manuel Villarreal"
date: '08/28/24'
output:
  xaringan::moon_reader:
    self_contained: true
    css: ['xaringan-themer.css', 'set-colors.css']
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.width = 7,
  fig.height = 5.8,
  fig.align = 'center',
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, echo=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#005f86",
  secondary_color = "#f8971f",
  inverse_header_color = "#FFFFFF"
)
```

```{r xaringan-extra, echo=FALSE, warning=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ))
rmarkdown::html_dependency_font_awesome()
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE)
```

```{r data-sim-example, echo = FALSE}
library(flipbookr)
set.seed(291792)
# simulating data for experiment one
blood <- data.frame("id" = seq(from = 1, to = 200),
                    "sex" = rep(x = c("female", "male"), each = 100),
                    "age" = round(
                      x = rnorm(n = 200, mean = 45, sd = 3) + 
                        rexp(n = 200, rate = 1/4),
                      digits = 0),
                    "height" = round(
                      x = rnorm(n = 200, 
                                mean = rep(x = c(170.7, 179.8), each = 100),
                                sd = rep(x = c(6.35, 7.62), each = 100)), 
                      digits = 0)) |>
  dplyr::mutate("blood_pressure" = round(x = 
                  96.6 + 16 * as.numeric(sex == "male") + 
                  0.7 * age - 0.4 * age * as.numeric(sex == "male") +
                  rnorm(n = 200, mean = 0, sd = 3), digits = 1))

n <- dim(blood)[1]
y_bar <- mean(blood$blood_pressure)
x_bar <- mean(blood$age)
s_xy <- sum(x = (blood$blood_pressure - y_bar) * (blood$age - x_bar))
s_xx <- sum(x = (blood$age - x_bar)^2)
hat_beta1 <- s_xy / s_xx
hat_beta0 <- y_bar - hat_beta1 * x_bar
hat_mu <- hat_beta0 + hat_beta1 * blood$age
sse <- sum((blood$blood_pressure - hat_mu)^2)
hat_sigma2 <- 1 / (n - 2) * sse
```

### OLS Estimators

- From last class we know that the OLS estimators are the solution to our 
decision problem in which we want to find candidate values for $\beta^*_0$ 
and $\beta^*_1$ that minimize the distance between the "predicted" line 
and our observations.

--

- In our blood pressure example, that means finding an estimate of the blood 
pressure of a person that is "0 years old" ( $\beta_0$ ).

--

- Finding an estimate of the difference in blood pressure for people who are 
one year of age apart ( $\beta_1$ ).

--

$$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}_1$$ $$\hat{\beta}_1 =  \frac{\sum_{i = 1}^n(y_i - \bar{y})(x_{i1} - \bar{x}_1)}{\sum_{i = 1}^n(x_{i1}-\bar{x}_1)^2}$$

---

### Assumptions

In classical linear regression there are 5 assumptions that allow us to make
inferences about the values of the parameters in the model using the "classical"
approach without worries.

--

1. Errors are centered around 0: $\mathrm{E}(\epsilon_i) = 0$.
--

1. Constant variance: $\mathrm{V}ar(\epsilon_i) = \sigma^2\ \text{for all}\ i = 1, 2, \dots, n$
--

1. Independence of errors: $\epsilon_1, \epsilon_2, \dots, \epsilon_n$.
--

1. Identically distributed errors: $\epsilon_i \sim \left(0, \sigma^2\right)$

--

Additionally, for now we will assume:

- Normally distributed errors: $\epsilon_i \overset{iid}{\sim} Normal\left(0, \sigma^2\right)$

---
### What does this mean?

Some consequences of these assumptions are that:

--

1. $\mathrm{E} \left(Y_i \mid X_i \right) = \beta_0 + \beta_1X_{i}$
--

1. $\mathrm{V}ar\left(Y_i \mid X_i \right) = \mathrm{V}ar\left(\epsilon_i\right) = \sigma^2$
--

1. The OLS estimators $\hat{\beta}_0$ and $\hat{\beta}_1$ are optimal.

--

Now, if we have 
$\epsilon_i \overset{iid}{\sim} Normal\left(0, \sigma^2\right)$
Then we know that:
$$\hat{\beta}_0 \sim N\left(\beta_0,\ \frac{\sigma^2\sum_{i=1}^n x_i^2}{n\sum_{i=1}^n (x_i - \bar{x})^2}\right)\quad,\quad \hat{\beta}_1 \sim N\left(\beta_1,\ \frac{\sigma^2}{\sum_{i=1}^n (x_i - \bar{x})^2}\right)$$

---
### What happens if we loose the normality assumption?

As long as our errors $\epsilon_i$ have an expectation equal to 0 
(**assumption 1**), constant variance (**assumption 2**), are 
independent (**assumption 3**), and are identically distributed 
(**assumption 4**).

--

Then by the Central Limit Theorem (**aka CLT**) we have that:
$$\hat{\beta}_0 \overset{\bullet}{\sim} N\left(\beta_0,\ \frac{\sigma^2\sum_{i=1}^n x_i^2}{n\sum_{i=1}^n (x_i - \bar{x})^2}\right)\quad,\quad \hat{\beta}_1 \overset{\bullet}{\sim} N\left(\beta_1,\ \frac{\sigma^2}{\sum_{i=1}^n (x_i - \bar{x})^2}\right)$$

--

- This means that our estimates follow **approximately** a Normal 
distribution. In other words, although our confidence intervals might not have
the correct length or our **p-values** will not be exactly right, they will 
be **pretty close!**

---
class: middle, center, inverse

### Let's do a simulation study!

Open a new R file on your project and we will do this one together.

---
### How do we test the model assumptions?

- The short answer is... we can't, however, we can use visualization methods 
that will indicate to us if something is wrong.

--

- First we will need to calculate the model errors which can be obtained with
the following function:
$$\hat{\epsilon}_i = y_i - \hat{\mu}_i$$

--

- These errors are known as **residuals** and we can use them to visually 
inspect if the assumptions of the model are "correct".

--

- We will go back to our blood pressure example, were we defined the model:
$$\text{blood pressure}_i = \beta_0 + \beta_1\text{age}_i + \epsilon_i \quad and \quad \hat{\mu}_i = \hat{\beta}_0 + \hat{\beta}_1 \text{age}_i$$

---
### Resdual plots

- The histogram should be approximately centered at zero and symmetric.

--

- A qq-plot is a particular class of scatter plots which takes the sampled 
quantiles of our observations and compares them to the theoretical quantiles 
of a distribution, in this case the standard normal.

--

- We use scatter plots to compare the residuals across different independent 
variables, for example the observation number (order).

--

- Auto-correlation plots to check that there is no correlation between our 
variables across time, for example that the residual $n$ is not  correlated to 
$n+1$.

--

- All these methods could allow us to evaluate if the **simple linear model** is 
a good approach to the data that we have.

---
### Residual histogram

.pull-left[
```{r residual-hist, eval = FALSE}
# calculate the expected valueK hat_mu 
hat_mu <- hat_beta0 + hat_beta1 * blood$age

# calculate the residuals
hat_epsilon <- blood$blood_pressure - hat_mu

# Then we can draw the histogram
hist(x = hat_epsilon, breaks = n/10,
     axes = FALSE, ann = FALSE, freq = FALSE,
     border = "white", col = "#f8971f")
abline(v = mean(hat_epsilon), lty = 2, lwd = 3,
       col = "#005f86")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
mtext(side = 1, cex = 2, line = 3,
      text = expression(
        x = paste("Residuals: ",
                  hat(epsilon)[i])))
curve(dnorm(x, mean = 0, sd = sqrt(hat_sigma2)),
      from = -12, to = 12, n = 100, add = TRUE)
```
]

.pull-right[
```{r residual-hist-out, ref.label = "residual-hist", fig.align = 'center', echo = FALSE, fig.height = 6}

```
]

---
### Residual qq-plot

.pull-left[
```{r residual-qq, eval = FALSE}
# draw a quantile-quantile plot 
# using the normal distribution
qqnorm(y = hat_epsilon, pch = 21, 
       col = "white", bg = "#f8971f",
       cex = 2, las = 1, axes = FALSE,
       ann = FALSE)

# add a comparison line
qqline(y = hat_epsilon, lwd = 3, lty = 2,
       col = "#005f86")

# editing plot
box(bty = "l")
axis(side = 1, cex.axis = 1.2)
axis(side = 2, las = 2, cex.axis = 1.2)
mtext(text = "Threoretical quantiles",
      line = 3, cex = 2, side = 1)
mtext(text = "Sample quantiles",
      line = 2.5, cex = 2, side = 2)
```
]

.pull-right[
```{r residual-qq-out, ref.label = "residual-qq", fig.align = 'center', echo = FALSE, fig.height = 6}

```
]

---
### Residual scatter plot (vs order)

.pull-left[
```{r residual-order, eval = FALSE}
# Plot residual against
# its corresponding id (order)
plot(x = blood$id, y = hat_epsilon, ann = FALSE, 
     axes = FALSE, cex = 2, pch = 21,
     col = "white", bg = "#f8971f")
box(bty = "l")
axis(side = 1, cex.axis = 1.5)
axis(side = 2, cex.axis = 1.5, las = 2)
mtext(text = "Residuals", side = 2, line = 2.5,
      cex = 2)
mtext(text = "Id", side = 1, line = 3,
      cex = 2)

# what should we expect?
abline(h = 0, col = "#005f86", lty = 4, 
       lwd = 3.5)
```
]

.pull-right[
```{r residual-order-out, ref.label = "residual-order", fig.align = 'center', echo = FALSE, fig.height = 6}

```
]

---
### Residual scatter plot (vs sex)

.pull-left[
```{r residual-sex, eval = FALSE}
# transform sex to a numeric variable
sex_id <- as.numeric(x = blood$sex == "female")

# Plot residual against the new variable
plot(x = jitter(x = sex_id, amount = 0.4),
  y = hat_epsilon, ann = FALSE, axes = FALSE,
  cex = 2, pch = 21, col = "white", 
  bg = ifelse(test = blood$sex == "female", 
              yes = "#f8971f", no = "#005f86"))
box(bty = "l")
axis(side = 1, cex.axis = 1.5, at = c(0, 1),
     labels = c("male", "female"))
axis(side = 2, cex.axis = 1.5, las = 2)
mtext(text = "Residuals", side = 2, line = 2.5,
      cex = 2)
mtext(text = "Sex", side = 1, line = 3, cex = 2)

# what should we expect?
abline(h = 0, col = "#000000", lty = 4, lwd = 4)
```
]

.pull-right[
```{r residual-sex-out, ref.label = "residual-sex", fig.align = 'center', echo = FALSE, fig.height = 6}

```
]

---
### Residual's auto-correlation

.pull-left[
```{r residual-autocor, eval = FALSE}
# We can use the function built in r
acf(x = hat_epsilon, las = 1, lwd = 2,
    main = "Series: residuals", 
    ylab = "", 
    xlab = "Lag", cex.lab = 1.7, cex.main = 1.5,
    cex.axis = 1.3)
mtext(text = "Auto-correlation", side = 2, 
      line = 2.8, cex = 1.7)
```

- **Note:** The auto-correlation values rarely go under the line, this 
means that there is a high level of autocorrelation across our observations at
different "time" (in this case order) lags.
]

.pull-right[
```{r residual-autocor-out, ref.label = "residual-autocor", fig.align = 'center', echo = FALSE, fig.height = 6}

```
]
