---
title: "How to do your homework"
subtitle: "Extra: Homeworks"  
author: 
  - "Manuel Villarreal"
date: '08/28/24'
output:
  xaringan::moon_reader:
    self_contained: true
    css: ['xaringan-themer.css', 'set-colors.css']
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.width = 7,
  fig.height = 5.8,
  fig.align = 'center',
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, echo=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#005f86",
  secondary_color = "#f8971f",
  inverse_header_color = "#FFFFFF"
)
```

```{r xaringan-extra, echo=FALSE, warning=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ))
rmarkdown::html_dependency_font_awesome()
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE)
```

### Homeworks

- Your first homework should be (or will be) available on the linear regression
module on CANVAS.

--

- As you will see there is a propmpt with an experimental question and some 
details about the design.

--

- This is all part of the information you need to do your homework.

--

- But what do you need to have on the pdf that you will turn in?

---
### Introduction

- When introducing the results section of o a paper it is always useful to 
remind people what it is that we are interested in.

--

- This should fit in a single paragraph and explain what the research question 
was and the type of variables or conditions that have been colected, for example
what are the experimental conditions.

---
### Sample

- A second step is to introduce details about the sample, what are the means 
of the variables of interest, how many participants are in each experimental 
condition etc.

--

- This section should also be short but give the reader a general idea of the 
structure of the study and some particularitues of the sample in it.

---
### Modeling

- Describe the modeling approach that would answer the research question, and 
if the model that you found to be the best is the same one then there is nothing 
else to do.

--

- On the other hand, if the model that you used is different from the simplest
model that would answer the experimental question you can explain your reasons
briefly.

--

- For example, is there an important reason to add one variable? does the 
additional variable make the model assumptions more plausible?

---
### Estimation

- Present the results fomr the model you choose as the best.

--

- Here it is important not to just throw numbers at the reader but to interpret
what does numbers mean and how do they relate to the experimental question.

--

- Remember to use in-line code to report estimated values and not to add more 
than 3 decimal places in the numbers you report.

--

- Adding figures when describing results is always better than just writing 
values on a table. As a rule of thumb if you can replace a table or in-line 
coding with a graph it is always better to use the graph and explain the results
with it.

---
### Finish

- Finish with a small conclusion of what the results mean, for example, how does
all the work that you've done answer the original research question?

--

- That should be all. Try not to make it more than 2 pages long including 
figures.

--

- It's very important to summarize information.