---
title: "In class activity 05"
output: pdf_document
date: "2024-09-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r data-sim-example, echo = FALSE}
set.seed(291792)
# simulating data for experiment one
blood <- data.frame("id" = seq(from = 1, to = 200),
                    "sex" = rep(x = c("female", "male"), each = 100),
                    "age" = round(
                      x = rnorm(n = 200, mean = 45, sd = 3) + 
                        rexp(n = 200, rate = 1/4),
                      digits = 0),
                    "height" = round(
                      x = rnorm(n = 200, 
                                mean = rep(x = c(170.7, 179.8), each = 100),
                                sd = rep(x = c(6.35, 7.62), each = 100)), 
                      digits = 0)) |>
  dplyr::mutate("blood_pressure" = round(x = 
                  96.6 + 16 * as.numeric(sex == "male") + 
                  0.7 * age - 0.4 * age * as.numeric(sex == "male") +
                  rnorm(n = 200, mean = 0, sd = 3), digits = 1))
```

1. Using data from the blood pressure example run two models using the `lm()` 
function, one that includes age and sex as covariates and one that includes 
age, sex, and the interaction.

```{r}
lm_1 <- lm(formula = blood_pressure ~ age + sex, data = blood)
lm_2 <- lm(formula = blood_pressure ~ age + sex + age * sex, data = blood)
```

1. Calculate the BIC for both models and compare them.
```{r}
bic_age_sex <- dim(blood)[1] * log(x = mean(lm_1$residuals ^ 2)) +
  4 * log(x = dim(blood)[1])

bic_int <- dim(blood)[1] * log(x = mean(lm_2$residuals ^ 2)) +
  5 * log(x = dim(blood)[1])

print(c(bic_age_sex,bic_int))
```

1. Write a conclusion stating which model you would choose from the two and 
interpret the values of the parameters.

1. Fit a model that adds height as a predictor (with age, sex, and the 
interactions). Calculate the BIC, is this new model better than the one that
doesn't include height?

```{r}
lm_3 <- lm(formula = blood_pressure ~ age + sex + height + age * sex, 
           data = blood)

bic_int_height <- dim(blood)[1] * log(x = mean(lm_3$residuals ^ 2)) +
  6 * log(x = dim(blood)[1])

print(c(bic_age_sex,bic_int, bic_int_height))
```



